Optimal Finite Element Integration

### 2015-08-04 TOMS_Paper/assembly.tex
Update hyperelasticity + minor fixes

\begin{abstract}
...
\end{abstract}

@@

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
\item We implemented the theory in our compiler, COFFEE\footnote{...}, which is the default optimization system in the Firedrake framework. Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
\item We provide an unprecedented (for extension) performance evaluation comparing several code transformation/synthesizers systems in a wide range of forms of increasing complexity, which is essential to demonstrate our optimality claim.
\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we deal with (i.e., optimize) them
\item We enrich the set of low-level transformations described in~\citeN{Luporini}
\end{itemize}

@@

\paragraph{Elasticity}
%\begin{figure}[t]  
%...
%%\includegraphics[scale=0.7]{perf-results/elasticity}
%%\caption{Elasticity results.}\label{fig:elasticity}
%\end{figure}
%Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem. This form uses a vector-valued space for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance. The \texttt{ffc} set of optimizations leads to notable improvements over the original code at $q=1$. The use of inderection arrays allows to phisically eliminate zero-valued columns at code generation time; as a consequence, different tabulated basis functions are merged into a single array. Therefore, despite the execution being purely scalar because of indirection arrays, the reduction in arithmetic intensity and register pressure imply improvement in performance. Nevertheless, \texttt{auto} remains in general the best choice, with gains over \texttt{ffc} that are wider as $p$ and $nf$ increase. 
%
%For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceed 1$\times$. On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$. 
%
%It is worth noting that speed ups of \texttt{auto} over \texttt{fix} decrease at $q=4$, particularly for low values of $p$. As we will discuss in Section~\ref{sec:auto-analysis}, this is because at $q=4$ the vector-register tiling transformation (in combination with loop unroll-and-jam) leads to the highest performance. In principle, vector-register tiling can be used in combination to the zero-avoidance technique; however, due to mere technical limitations, this is currently not supported in COFFEE. Once solved, we expect much higher speed ups in the $q=4$ regime as well.

\paragraph{Hyperelasticity}
\begin{figure}[t]
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a Hyperelastic model}\label{fig:helmholtz}
\end{figure}

%
%Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 2$ could not be executed because of FEniCS-Form-Compiler's technical limitations. 
%
%For \texttt{auto}, massive speed ups for $q \geq 2$ are to be ascribed to aggressive and successful expression writing. Hyperelasticity problems are really compute-intensive, with thousands of operations being performed, so reductions in redundant and useless computation are crucial. Complex forms like hyperelasticity would benefit from further ``specialized'' optimizations: for example, it is a known technical limitation of COFFEE that, in some circumstances, less temporaries could (should) be generated and that hoisted code could (should) be suitably distributed over different loops to minimize register pressure (e.g. COFFEE could apply loop fission for obtaining significantly better register usage). We expect to obtain considerably faster code once such optimizations will be incorporated. 
%
%In the regime $q \geq 2$ and $nf=1$, peformance improvements are less pronounced moving from $p=1$ to $p=2$, although still significant; in particular, we notice a drop at $p=2$, followed by a raise up to $p=4$. It is worth observing that this effect is common to all sets of optimizations. The hypothesis is that this is due to the way coefficient functions are evaluated at quadrature points (identical in all configurations), which cannot be easily vectorized unless a change in storage layout and loops order is implemented in the code (abstract syntax tree) generator on top of COFFEE.

>>>>

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to compute a result. We establish a model under which optimality can systematically be reached and comment on the general applicability of this model. The model is integrated in a compiler such that, given a representation of an integration routine, a loop nest in optimal form can automatically be derived. Extensive performance evaluation proves the robustness of the model, showing considerable performance improvements over a number of alternative code generation systems. The effect of low-level optimization is also discussed.
\end{abstract}

@@

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
\item We implemented the theory in our compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is the default optimization system in the Firedrake framework. Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
\item We provide an unprecedented (for extension) performance evaluation comparing several code transformation/synthesizers systems in a wide range of forms of increasing complexity, which is essential to demonstrate our optimality claim.
\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we deal with (i.e., optimize) them
\item We enrich the set of low-level transformations described in~\citeN{Luporini}
\end{itemize}

@@

\paragraph{Elasticity}
%\begin{figure}[t]  
%...
%%\includegraphics[scale=0.7]{perf-results/elasticity}
%%\caption{Elasticity results.}\label{fig:elasticity}
%\end{figure}

\paragraph{Hyperelasticity}
\begin{figure}[t]
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a Hyperelastic model}\label{fig:hyperelasticity}
\end{figure}
In the hyperelastic model experiments, pre-evaluation is never triggered by \texttt{cfO2}. This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction. Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques to this purpose. Results are displayed in Figure~\ref{fig:hyperelasticity}. \texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions. \texttt{ufls} performs generally well and exhibits significant speed ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.
